---
title: "Choropleth maps with tricolore"
author: "Jonas Schöley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choropleth maps with tricolore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{shiny, sf, leaflet, tricolore, dplyr, ggplot2, ggtern, httpuv}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  tidy = FALSE,
  comment = "#>",
  fig.width = 6, fig.height = 6
)
```

Here I demonstrate how to use the `tricolore` library to generate ternary choropleth maps using both `ggplot2` and `leaflet`.

The data
--------

```{r}
library(tricolore)
```

The data set `euro_example` contains the administrative boundaries for the European NUTS-2 regions in the column `geometry`. This data can be used to plot a choropleth map of Europe using the `sf` package. Each region is represented by a single row. The name of a region is given by the variable `name` while the respective [NUTS-2](https://en.wikipedia.org/wiki/Nomenclature_of_Territorial_Units_for_Statistics) geocode is given by the variable `id`. For each region some compositional statistics are available: Variables starting with `ed` refer to the relative share of population ages 25 to 64 by educational attainment in 2016 and variables starting with `lf` refer to the relative share of workers by labor-force sector in the European NUTS-2 regions 2016.

Take the first row of the data set as an example: in the Austrian region of "Burgenland" (`id` = `AT11`) 16.5% of the population aged 25--64 had attained an education of "Lower secondary or less" (`ed_0to2`), 55.7% attained "upper secondary" education (`ed_3to4`), and 27.9% attained "tertiary" education. In the very same region 4.4% of the labor-force works in the primary sector, 26.8% in the secondary and 68.2% in the tertiary sector.

The education and labor-force compositions are *ternary*, i.e. made up from three elements, and therefore can be color-coded as the weighted mixture of three primary colors, each primary mapped to one of the three elements. Such a color scale is called a *ternary balance scheme*^[See for example Dorling (2012) and Brewer (1994).]. This is what `tricolore` does.

`ggplot2` for ternary choropleth maps
-------------------------------------

Here I show how to create a choropleth map of the regional distribution of education attainment in Europe 2016 using `ggplot2`.

**1. Using the `Tricolore()` function, color-code each educational composition in the `euro_example` data set and add the resulting vector of hex-srgb colors as a new variable to the data frame. Store the color key seperately.**

```{r}
# color-code the data set and generate a color-key
tric <- Tricolore(euro_example, p1 = 'ed_0to2', p2 = 'ed_3to4', p3 = 'ed_5to8')
```

`tric` contains both a vector of color-coded compositions (`tric$rgb`) and the corresponding color key (`tric$key`). We add the vector of colors to the map-data.

```{r}
# add the vector of colors to the `euro_example` data
euro_example$rgb <- tric$rgb
```

**2. Using `ggplot2` and the joined color-coded education data and geodata, plot a ternary choropleth map of education attainment in the European regions. Add the color key to the map.**

The secret ingredient is `scale_fill_identity()` to make sure that each region is colored according to the value in the `rgb` variable of `euro_educ_map`.

```{r}
library(ggplot2)

plot_educ <-
  # using sf dataframe `euro_example`...
  ggplot(euro_example) +
  # ...draw a polygon for each region...
  geom_sf(aes(fill = rgb, geometry = geometry), size = 0.1) +
  # ...and color each region according to the color code in the variable `rgb`
  scale_fill_identity()

plot_educ 
```

Using `annotation_custom()` and `ggplotGrob` we can add the color key produced by `Tricolore()` to the map. Internally, the color key is produced with the [`ggtern`](http://www.ggtern.com/) package. In order for it to render correctly we need to load `ggtern` *after* loading `ggplot2`. Don't worry, the `ggplot2` functions still work.

```{r}
library(ggtern)
plot_educ +
  annotation_custom(
    ggplotGrob(tric$key),
    xmin = 55e5, xmax = 75e5, ymin = 8e5, ymax = 80e5
  )
```

Because the color key behaves just like a `ggplot2` plot we can change it to our liking.

```{r}
plot_educ <-
  plot_educ +
  annotation_custom(
    ggplotGrob(tric$key +
                 theme(plot.background = element_rect(fill = NA, color = NA)) +
                 labs(L = '0-2', T = '3-4', R = '5-8')),
    xmin = 55e5, xmax = 75e5, ymin = 8e5, ymax = 80e5
  )
plot_educ
```

Some final touches...

```{r}
plot_educ +
  theme_void() +
  coord_sf(datum = NA) +
  labs(
   title = 'European inequalities in educational attainment',
      subtitle = 'Regional distribution of ISCED education levels for people aged 25-64 in 2016.'
  )
```

`leaflet` for ternary choropleth maps
-------------------------------------

The `ggplot2` example above is easily adapted to `leaflet`. This time I use a continuous color scale.

```{r}
# color-code the data set and generate a color-key
tric <- Tricolore(euro_example, p1 = 'ed_0to2', p2 = 'ed_3to4', p3 = 'ed_5to8',
                  breaks = Inf)

# add the vector of colors to the `euro_example` data
euro_example$rgb <- tric$rgb
```

`leaflet` requires geodata in spherical coordinates (longitude-latitude format). Therefore I reproject the data to a [suitable crs](http://spatialreference.org/ref/epsg/wgs-84/) using the `sf` package.

```{r}
library(sf)
library(leaflet)

euro_example %>%
  st_transform(crs = 4326) %>%
  leaflet() %>%
  addPolygons(smoothFactor = 0.1, weight = 0,
              fillColor = euro_example$rgb,
              fillOpacity = 1)
```

Adding a background map gives geographical context to the map. I also add a mouse pop-up of the actual data.

```{r}
euro_example %>%
  st_transform(crs = 4326) %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldTerrain) %>%
  addPolygons(smoothFactor = 0.1, weight = 0,
              fillColor = euro_example$rgb,
              fillOpacity = 1,
              popup =
                paste0(
                  '<b>', euro_example$name, '</b></br>',
                  'Primary: ',
                  formatC(euro_example$ed_0to2*100,
                          digits = 1, format = 'f'), '%</br>',
                  'Secondary: ',
                  formatC(euro_example$ed_3to4*100,
                          digits = 1, format = 'f'), '%</br>',
                  'Tertiary: ',
                  formatC(euro_example$ed_5to8*100,
                          digits = 1, format = 'f'), '%</br>'
                )
  )
```

Adding the legend to the leaflet map requires a bit of a [hack](https://github.com/rstudio/leaflet/issues/51#issuecomment-213108125).

```{r}
makePlotURI <- function(expr, width, height, ...) {
  pngFile <- shiny::plotPNG(function() { expr }, width = width, height = height, ...)
  on.exit(unlink(pngFile))

  base64 <- httpuv::rawToBase64(readBin(pngFile, raw(1), file.size(pngFile)))
  paste0("data:image/png;base64,", base64)
}

legend_symbol <- makePlotURI({
  print(tric$key +
          theme(plot.background = element_rect(fill = NA, color = NA)) +
          labs(L = '0-2', T = '3-4', R = '5-8'))
}, 200, 200, bg = "transparent")

df <- data.frame(
  lng = 30,
  lat = 70,
  plot = legend_symbol,
  stringsAsFactors = FALSE
)

euro_example %>%
  st_transform(crs = 4326) %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(smoothFactor = 0.1, weight = 0,
              fillColor = euro_example$rgb,
              fillOpacity = 1,
              popup =
                paste0(
                  '<b>', euro_example$name, '</b></br>',
                  'Primary: ',
                  formatC(euro_example$ed_0to2*100,
                          digits = 1, format = 'f'), '%</br>',
                  'Secondary: ',
                  formatC(euro_example$ed_3to4*100,
                          digits = 1, format = 'f'), '%</br>',
                  'Tertiary: ',
                  formatC(euro_example$ed_5to8*100,
                          digits = 1, format = 'f'), '%</br>'
                )
  ) %>%
  addMarkers(data = df, icon = ~icons(plot))
```

Literature
----------

Brewer, C. A. (1994). Color Use Guidelines for Mapping and Visualization. In A. M. MacEachren & D. R. F. Taylor (Eds.), Visualization in Modern Cartography (pp. 123–147). Oxford, UK: Pergamon.

Dorling, D. (2012). The Visualization of Spatial Social Structure. Chichester, UK: Wiley.
